---
title: "Statistical Rethinking"
subtitle: Chapter 6
output:
  html_document:
    df_print: paged
---

```{r}
library(rethinking)
library(tidyverse)
library(reshape2)
```

### Practice

#### Medium

##### Exercise 1


Modify the DAG on page 190 to include the variable $V$, an unobserved cause of $C$ and $Y$: 
\[
C \leftarrow V \rightarrow Y.
\]
Reanalyze the DAG. How many paths connect $X$ to $Y$? Which must be closed? Which variables should you condition on now?

---

```{r}
library(dagitty)

dag_6.1 <- dagitty("dag{
  U [unobserved]
  V [unobserved]
  X -> Y
  X <- U <- A -> C -> Y 
  U -> B <- C
  C <- V -> Y
  }")

adjustmentSets(dag_6.1, exposure="X", outcome="Y")
```

- There are four paths now!

- There are two open paths that need to be closed.

- it is sufficient to close A.

#### Hard

##### Exercise 1

Use the Waffle House data to find the total causal influence of number of Waffle Houses on divorce rate. Justify your model or models with a causal graph.

---

```{r}
data("WaffleDivorce")
WaffleDivorce
```

```{r}
library(dagitty)
dag_6.2 <- dagitty( "dag {
  
  A -> D 
  A -> M -> D
  A <- S -> M
  S -> W -> D}"
  )
adjustmentSets(dag_6.2, exposure="W", outcome="D")
```

```{r}
d <- WaffleDivorce
d2 <- list()

d2$W <- standardize(d$WaffleHouses)
d2$D <- standardize(d$Divorce)
d2$S <- d$South + 1

model <- quap(alist(D ~ dnorm(mu, sigma),
                   mu <- a[S] + bW*W,
                   a[S] ~ dnorm(0, 1),
                   bW ~ dnorm(0, 2),
                   sigma ~ dexp(1)), 
             data=d2) 

precis(model, depth=2)
```





